Predicting Conflict Escalation

### 2023-05-04 docs/hes_final/article.qmd
add test accuracy maps, use markdown tables

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
    margin-left: 25mm
    margin-right: 25mm
    margin-top: 25mm
    margin-bottom: 25mm
    whitespace: medium
    mermaid:
      theme: neutral
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
      \setlength{\belowcaptionskip}{-5pt}
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
  The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources. The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).

@@

::: {#fig-transformer layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 3
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Spike -->|included during model training| Trans1
```
Transformer model architecture
:::
## Results

We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc, @fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results. Each of the models outperformed the other two in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics. Two models using country features seem to perform
better on these two metrics.

@@

![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){#fig-best-max-metrics}

![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){#fig-best-min-metrics}

@@

@@

![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){#fig-ffnn-test-predictions}

![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){#fig-xgboost-test-predictions}

![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){#fig-transformer-test-predictions}

@@

We think that all three approaches produced promising results and the model choice depends on the developers' goals, performance metrics and available resources.

>>>>

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
    margin-left: 25mm
    margin-right: 25mm
    margin-top: 25mm
    margin-bottom: 25mm
    whitespace: medium
    mermaid:
      theme: neutral
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
       \usepackage[font=small,labelfont={bf,small}]{caption}
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
      \setlength{\belowcaptionskip}{-5pt}
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
  The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of over 2.5 million document embeddings derived from publicly available news sources. The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).

@@

::: {#fig-transformer layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Spike -->|included during model training| Trans1
```
Transformer model architecture
:::

## Results

We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results. Each of the models outperformed the other two in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics. Two models using country features seem to perform better on these two metrics.

@@

\onecolumn

::: {layout="[[1, 1]]"}
| metric               | model       |   value |
|:---------------------|:------------|--------:|
| False Discovery Rate | Transformer |    0.80 |
| False Discovery Rate | XGBoost     |    0.83 |
| False Discovery Rate | FFNN        |    0.74 |
| False Negative Rate  | Transformer |    0.53 |
| False Negative Rate  | XGBoost     |    0.28 |
| False Negative Rate  | FFNN        |    0.59 |
| False Positive Rate  | Transformer |    0.30 |
| False Positive Rate  | XGBoost     |    0.51 |
| False Positive Rate  | FFNN        |    0.27 |

: Model performance metrics (smaller values are better) {#tbl-best-min-metrics}

| metric             | model       |   value |
|:-------------------|:------------|--------:|
| Cohen Kappa Score  | XGBoost     |    0.09 |
| Cohen Kappa Score  | FFNN        |    0.12 |
| Cohen Kappa Score  | Transformer |    0.11 |
| F1 Score           | FFNN        |    0.32 |
| F1 Score           | XGBoost     |    0.28 |
| F1 Score           | Transformer |    0.28 |
| Precision Score    | XGBoost     |    0.17 |
| Precision Score    | FFNN        |    0.26 |
| Precision Score    | Transformer |    0.20 |
| Recall Score       | XGBoost     |    0.72 |
| Recall Score       | Transformer |    0.47 |
| Recall Score       | FFNN        |    0.41 |
| Roc Auc Score      | Transformer |    0.64 |
| Roc Auc Score      | XGBoost     |    0.69 |
| Roc Auc Score      | FFNN        |    0.64 |
| True Negative Rate | Transformer |    0.70 |
| True Negative Rate | XGBoost     |    0.49 |
| True Negative Rate | FFNN        |    0.73 |
| True Positive Rate | FFNN        |    0.41 |
| True Positive Rate | XGBoost     |    0.72 |
| True Positive Rate | Transformer |    0.47 |
: Model performance metrics (larger values are better) {#tbl-best-max-metrics}

:::
\twocolumn

\onecolumn

::: {layout="[[1, 1]]"}

@@

:::
::: {layout="[[1, 1]]"}

![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy}

![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy}

:::
::: {layout="[[1, 1]]"}

![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}

:::
\twocolumn

@@

@@

We think that all three approaches produced promising results and the model choice depends on the user's goals, performance metrics and available resources.
